NDNS multilevel LCA CW2CB2 or CW2CB3

### 2018-09-01 paper/Chapters/Chapter2.tex
grammarly checked

Participants were asked to keep a record of everything eaten or drunk over four consecutive days. Interviewers undertook three visits with each participant. At the first visit, the interviewer explained the method followed a protocol, taking participants through the sections in the diary including how to describe details of food and drink and portion size and an example day. The second was a brief visit to check for compliance, answer questions or deal with problems and review the diary to identify and edit possible omissions and missing detail. The third visit was to collect the diary and again review and edit possible omissions.

In the diary, participants were asked to record portion sizes in household measures (e.g. one tablespoon of beans, one Kit Kat finger-size), or for packaged foods to note the weight indicated on the packet. For homemade dishes, participants were asked to record on a separate page in the diary the individual ingredients and quantities for the whole dish along with a brief description of the cooking method and how much of dish they had consumed. For each eating occasion, in addition to the details of what and how much was eaten, participants were also asked to record: when was it, where they were, and who they were eating with. An example, used as guidance for participants, of a food diary for one day is shown in \textbf{Appendix \ref{AppendixE}}.\vspace{-0.3cm}

@@

\begin{itemize}
	\item Total energy intake = (protein $\times$ 17) + (fat $\times$ 37) + (carbohydrate $\times$ 16) + (alcohol $\times$ 29)  kJ;
	\item Carbohydrate intake = total sugars + starch; 
	\item Where all quantities above were measured as mass in grams.
\end{itemize}

@@

Another two sets of weights were generated to correct for differential non-response (either due to refusal or inability) to 1) nurse visit and 2) giving a blood sample. Response rates to the nurse visit among those who completed a dietary diary was approximately 75\%, to the blood sample in adults were 51\%, 57\%, and 50\% for years 1 to 4, 5 to 6, and 7 to 8, respectively. In creating the nurse/blood sample weight, a logistic regression model was used by the NDNS RP study team to model the relationship between response to nurse visit/giving a blood sample and a set of predictor variables (socio-demographic, participant and catering/household unit characteristics). The model generated a predicted probability for each participant, which is the probability of agreeing to a nurse visit/providing a blood sample, given the characteristics of the individual and the household unit. Participants with characteristics associated with non-response were under-represented in the sample and therefore receive a low predicted probability. The inverses of these predicted probabilities were used as a set of non-response weights so that participants with a low predicted probability got a larger weight, increasing their representation in the sample. Then the nurse/blood sample weights were re-scaled so that the sum of the weights equalled the number of participants who had a nurse visit or who provided a blood sample. The final nurse/blood weights should, therefore, make the sample participants representative of all eligible persons in the population.

@@

Specifically, blood pressure was measured in a sitting position using an automated, validated machine, the Omron HEM907, after a five-minute rest. The mean of the second and third readings, taken at one-minute intervals, were used in the current report. Hypertension was defined as with systolic blood pressure of 140 mmHg or above, and/or diastolic blood pressure of 90 mmHg or above, and/or taking any medication specifically to reduce blood pressure.

@@

Blood samples were stored at 4 $^\circ$C, and sent directly by post to the Department of Haematology and Department of Clinical Biochemistry and Immunology, Addenbrooke's Hospital, Cambridge within two hours of their collection. Serum samples were obtained by centrifugation of the coagulated blood sample. Serum total, High-Density Lipoprotein (HDL) and Low-Density Lipoprotein (LDL) cholesterol, triglycerides (TG), fasting blood glucose, haemoglobin HbHbA1C were measured. HbA1C value of 6.5\% was used as the cut off point for diagnosing diabetes.

@@

\textbf{Equation \ref{LCA}} indicates that responses for an observation to the measuring variables are independent of one another given its membership in latent class $t$. However, in the NDNS RP data set, the assumption of independent observations is violated. Each individual completed their dietary diary for three/four consecutive days, their diary recordings were later converted into three/four vectors of categorical responses reflecting the type of carbohydrate consumption at each time slot of the day. The observed sequences (observed days) are nested within the participants and therefore are not independent. This nested data structure requires multilevel techniques. 
\vspace{-0.3cm}

@@

Proposed by Vermunt \parencite{Vermunt, vermunt2008latent} and Asparouhov and Muth\'en \parencite{muthen2009multilevel},  a traditional, parametric approach can be applied using a logistic regression model. For example, let's assume that there are two types of observation days in the dietary survey--high and low carbohydrate eating days. In an unconditional logistic regression model, the probability of the outcome (i.e. an observed high carbohydrate eating day vs. a low carbohydrate eating day) is constant within the individual-level, which means for each person throughout his/her survey there is some probability of following a high carbohydrate eating day. A random effect considers the individuals (level 2) to be drawn from the adult population in the UK, and the probability of the outcome (i.e. high carbohydrate eating days) across individuals is considered to be a random variable \parencite{snijders2011multilevel}.

@@

\begin{itemize}
	\item $\text{logit}{(x)} = \log{\frac{x}{1-x}}$, where $x \in [0,1]$;
	\item $P(C_{ij} = 1)$ as the probability that the randomly selected $i$th observation day of the $j$th individual is a high carbohydrate eating day;
	\item $\beta_{0j}$ as the random intercept, for the outcome $C_{ij} = 1$; 
	\item the random deviation of the individuals $u_{0j}$ are assumed be normally distributed (i.e. $u_{0j} \sim N(0, \sigma_{u_0}^2)$), the magnitude of the $u_{0j}$ variance ($\sigma_{u_0}^2$) indicates the influence of the individuals (level 2);
	\item $x_{ij}, w_j$ are the predictors for respectively day-level (weekdays or weekends) and individual-level, such as age, and sex.
\end{itemize}

@@

If the day-level LCA model (carbohydrate eating temporal pattern typologies) is best defined by $T (T \geqslant 2)$ latent classes, then $T-1$ random intercept will be specified by a two-level multinomial logistic regression model. Similar to the typical LCA models, the latent class variable in a MLCA is defined by multiple observed indicators (here is defined by the responses of eating carbohydrate within each time slots, throughout three/four consecutive days of their survey period). Considering the latent class indicators as indicator variables ($U_{ijk}$), the MLCA model can be written as follows:\vspace{-1cm}

@@

Since the parametric approach discussed above can be extremely computationally demanding \parencite{van2008using, vermunt2008latent}, an alternative approach is using a non-parametric MLCA \parencite{davidian2008growth}. In this approach, separate latent class models are specified for level 1 (observation days) and level 2 (individuals). Similar with the parametric MLCA approach, there are $T-1$ random intercepts, where $T$ is the number of level 1 latent classes. However, rather than assuming the random intercepts following a normal distribution, the non-parametric MLCA assumes a multinomial (discrete) distribution of the level 2 latent classes. This approach is less computationally demanding compared with the parametric approach. These level 2 (individual) latent classes reflect differences in the probability of belonging to a specific day-level latent class, so that individuals that have observation days with similar probabilities for the level 1 latent classes will be grouped together. The non-parametric MLCA model can be defined as follows:

@@

According to Finch and French's simulation study \parencite{finch2014multilevel}, non-parametric approach generally resulted in more accurate recovery of the underlying latent structure of the data at both levels and provided better latent class model compared with parametric approach. In the current project, we are interested in exploring both meaningful individual (level 2) latent classes and the daily (level 1) carbohydrate consumption classification. Therefore, non-parametric MLCA was employed 1) to identify latent classes of observation days (level 1) based on the subjects' responses to the 4-day food and drink diary and 2) to form distinct latent classes of individuals (level 2) based on the distribution of day-level carbohydrate eating temporal latent classes within individuals.\vspace{-0.3cm}

@@

The number of classes in level 1 was determined by 1) the evaluation of model fit indices, including the Bayesian information criterion (BIC) and entropy, which is a statistic that summarizes latent class probabilities where values near 1 indicate better latent class separation; 2) the Lo-Mendell-Rubin Likelihood Ratio Test (LMR-LRT) \parencite{lo2001testing, nylund2007deciding} which compares $q$ vs. $q-1$ classes models, where $q$ is the number of latent classes; most importantly, 3) pattern interpretability. In the steps of performing multilevel LCA, where LMR-LRT is not available, the same rules of model fit indices and pattern interpretability were used to determine the optimal combination of latent classes in observation day-level and individual-level. MLCA models were fitted in Mplus 7.4 \parencite{muthen2005mplus}, the Mplus syntax and outputs are shown in \textbf{Appendix \ref{AppendixB}}. \vspace{-0.3cm}

@@

Day-level latent classes identified by the first step of MLCA were tabulated according to the day of week and also whether the diary was recorded during weekends or not. A contingency table giving the frequency of responses across the 7-time slots of the survey days was produced. Descriptive statistics for the dietary day-level recordings according to the latent class memberships were presented. Pearson $\chi^2$ test was used to assess evidence fora difference in the distribution of categorical variables. One-way Analysis of Variance (ANOVA) was used to compare the means across the multiple groups for continuous variables.

@@

Associations between individual-level carbohydrate eating classes and hypertension (yes/no), body mass index (BMI, kg/m\textsuperscript{2}), and waist circumference (WC, cm) were explored in men and women separately. Point estimates of weighted means and proportions and 95\%CI of the characteristics were determined by applying either nurse visiting weights (for outcomes of hypertension, BMI, and WC) or blood sample weights (for diagnosis of DM) accordingly. Similarly, \textit{F} tests (for continuous variables) and adjusted Pearson $\chi^2$ tests (for categorical variables) were used to determine sex-specific differences by hypertension status, and BMI categories.

@@

\begin{enumerate}
	\item The crude association between the carbohydrate eating groups and the outcomes was first examined. 
	\item Potential confounders of the association between carbohydrate eating groups (exposure) and the outcomes were selected depending on the descriptive statistical analyses conducted above, i.e. those associated with both the exposure and the outcome and also not on the causal pathway were selected as potential confounders. Covariates that are strongly related with the outcomes but not associate with carbohydrate eating groups may reduce the standard errors and improve the precisions, so they were also considered in the linear regression models. 
	\item Confounding and/or interaction effects from each of the potential factors were checked one by one. Interaction effects were tested using the adjusted Wald test for whether the regression coefficients of the interaction terms are simultaneously equal to zero.  
	\item A preliminary model that includes all of the variables suggested to be confounders in the previous step was established. 
	\item The remaining variables were added to the preliminary model one by one to see if any of them may be a confounder in the presence of the other covariates. 
	\item For logistic regression models (hypertension) under the survey data, goodness-of-fit was assessed using \textit{F}-adjusted mean residual test \parencite{archer2001goodness,archer2007goodness} which is implemented in Stata using command \code{svylogitgof} \parencite{archer2006goodness}. Other diagnostics for regular logistic regression models, such as estimating the pseudo-R\textsuperscript{2},  AIC or BIC, checking the standardized Pearson residuals, or covariate pattern residuals are currently not available for weighted survey data. 
	\item For linear regression models (WC, BMI), the assumption of independent observations is violated because of the clustered structure of the data. Therefore, inference from such models is based on design rather than the model.  
	General checking such as QQ plots of the residuals (normality), plotting the residuals against fitted values (constant variance) are unneeded (also not available). Outliers, leverage, and Cook's distance cannot be checked either, however, participants with extreme weightings (if any) were checked by removing them and refitting the models as a sensitivity analysis.
	\item Since under complex survey designed data, the sampling-weighted least squares are not maximum likelihood, it would not be possible to compare models using likelihood ratio test. Instead, adjusted Wald tests with $p < 0.05$ were used as criteria for variable inclusion in the final model. Another Stata command \code{linktest} was used to decide whether quadratic and cubic terms of continuous variables were necessary in improving fitting of the model \parencite{pregibon1980goodness}. 
\end{enumerate}

>>>>

Participants were required to keep a diary that record everything eaten or drunk over four consecutive days. Interviewers undertook three visits with each participant. At the first visit, the interviewer explained the method followed a protocol, taking participants through the sections in the diary including how to describe details of food and drink and portion size and an example day. The second visit was a brief one to check for compliance, answer potential questions, and deal with problems. Interviewers also reviewed the diary to identify and edit possible omissions and missing detail in the food diary. The third visit was to collect the diary and again review and edit possible omissions.

In the diary, portion sizes of food or drink eaten were asked to be recorded in household measures (e.g. one tablespoon of beans, one Kit Kat finger-size), or for packaged foods to note the weight indicated on the packet. Homemade dishes were also described in the diary about their ingredients, quantities, along with a brief description of the cooking procedure and how much of the dish were consumed. For each eating occasion, in addition to the details of what and how much was eaten, participants were also asked to record: when was it, where they were, and who they were eating with. An example, used as guidance for participants, of a food diary for one day is shown in \textbf{Appendix \ref{AppendixE}}.\vspace{-0.3cm}

@@

\begin{itemize}
	\item Total energy intake = (protein $\times$ 17) + (fat $\times$ 37) + (carbohydrate $\times$ 16) + (alcohol $\times$ 29)  kJ;
	\item Carbohydrate intake = total sugars + starch; 
	\item All quantities above were measured as the mass in grams.
\end{itemize}

@@

Another two sets of weights were generated to correct for differential non-response (either due to refusal or inability) to 1) nurse visit and 2) giving a blood sample. Response rates to the nurse visit among those who completed a dietary diary was approximately 75\%, to the blood sample in adults were 51\%, 57\%, and 50\% for years 1 to 4, 5 to 6, and 7 to 8, respectively. In creating the nurse/blood sample weight, a logistic regression model was used by the NDNS RP study team to model the relationship between response to nurse visit/giving a blood sample and a set of predictor variables (socio-demographic, participant and catering/household unit characteristics). The model generated a predicted probability for each participant, which is the probability of agreeing to a nurse visit/providing a blood sample, given the characteristics of the individual and the household unit. Participants with characteristics associated with non-response were under-represented in the sample and therefore receive a low predicted probability. The inverses of these predicted probabilities were used as a set of non-response weights so that participants with a low predicted probability got a larger weight, increasing their representation in the sample. Then the nurse/blood sample weights were re-scaled so that the sum of which equalled the number of participants who had a nurse visit or who provided a blood sample. The final nurse/blood weights should, therefore, make the sample participants representative of all eligible persons in the population.

@@

Specifically, blood pressure was measured in a sitting position using an automated, validated machine, the Omron HEM907, after a five-minute rest. The mean of the second and third readings, taken at one-minute intervals, were used in the current report. Hypertension was defined as with systolic blood pressure of 140 mmHg or above, or diastolic blood pressure of 90 mmHg or above, or taking any medication specifically to reduce blood pressure.

@@

Blood samples were stored at 4 $^\circ$C, and sent directly by post to the Department of Haematology and Department of Clinical Biochemistry and Immunology, Addenbrooke's Hospital, Cambridge within two hours of their collection. Serum samples were collected by centrifugation of the coagulated blood sample. Serum total, High-Density Lipoprotein (HDL) and Low-Density Lipoprotein (LDL) cholesterol, triglycerides (TG), fasting blood glucose, haemoglobin HbHbA1C were measured. HbA1C value of 6.5\% was used as the cut off point for diagnosing diabetes.

@@

\textbf{Equation \ref{LCA}} indicates that responses for an observation to the measuring variables are independent of one another given its membership in latent class $t$. However, in the NDNS RP data set, the assumption of independent observations is violated. Each individual completed their dietary diary for three/four consecutive days, and their diary recordings were later converted into three/four vectors of categorical responses reflecting the type of carbohydrate consumption at each time slot of the day. The observed sequences (observed days) are nested within the participants and therefore are not independent. This nested data structure requires multilevel techniques. 
\vspace{-0.3cm}

@@

Proposed by Vermunt \parencite{Vermunt, vermunt2008latent} and Asparouhov and Muth\'en \parencite{muthen2009multilevel},  a traditional, parametric approach can be applied using a logistic regression model. For example, let's assume that there are two types of observation days in the dietary survey--high and low carbohydrate eating days. In an unconditional logistic regression model, the probability of the outcome (i.e. an observed high carbohydrate eating day vs a low carbohydrate eating day) is constant within the individual-level, which means for each person throughout his/her survey there is some probability of following a high carbohydrate eating day. A random effect considers the individuals (level 2) to be drawn from the adult population in the UK, and the probability of the outcome (i.e. high carbohydrate eating days) across individuals is considered to be a random variable \parencite{snijders2011multilevel}.

@@

\begin{itemize}
	\item $\text{logit}{(x)} = \log{\frac{x}{1-x}}$, where $x \in [0,1]$;
	\item $P(C_{ij} = 1)$ as the probability that the randomly selected $i$th observation day of the $j$th individual is a high carbohydrate eating day;
	\item $\beta_{0j}$ as the random intercept, for the outcome $C_{ij} = 1$; 
	\item the random deviation of the individuals $u_{0j}$ are assumed to be normally distributed (i.e. $u_{0j} \sim N(0, \sigma_{u_0}^2)$), the magnitude of the $u_{0j}$ variance ($\sigma_{u_0}^2$) indicates the influence of the individuals (level 2);
	\item $x_{ij}, w_j$ are the predictors for respectively day-level (weekdays or weekends) and individual-level, such as age, and sex.
\end{itemize}

@@

If the day-level LCA model (carbohydrate eating temporal pattern typologies) is best defined by $T (T \geqslant 2)$ latent classes, then $T-1$ random intercept will be specified by a two-level multinomial logistic regression model. Similar to the typical LCA models, the latent class variable in an MLCA is defined by multiple observed indicators (here is defined by the responses of eating carbohydrate within each time slots, throughout three/four consecutive days of their survey period). Considering the latent class indicators as indicator variables ($U_{ijk}$), the MLCA model can be written as follows:\vspace{-1cm}

@@

Since the parametric approach discussed above can be extremely computationally demanding \parencite{van2008using, vermunt2008latent}, an alternative approach is using a non-parametric MLCA \parencite{davidian2008growth}. In this approach, separate latent class models are specified for level 1 (observation days) and level 2 (individuals). Similar with the parametric MLCA approach, there are $T-1$ random intercepts, where $T$ is the number of level 1 latent classes. However, rather than assuming the random intercepts following a normal distribution, the non-parametric MLCA assumes a multinomial (discrete) distribution of the level 2 latent classes. This approach is less computationally demanding compared with the parametric approach. These level 2 (individual) latent classes reflect differences in the probability of belonging to a specific day-level latent class so that individuals that have observation days with similar probabilities for the level 1 latent classes will be grouped together. The non-parametric MLCA model can be defined as follows:

@@

According to Finch and French's simulation study \parencite{finch2014multilevel}, the non-parametric approach generally resulted in a more accurate recovery of the underlying latent structure of the data at both levels and provided better latent class model compared with the parametric approach. In the current project, we are interested in exploring both meaningful individual (level 2) latent classes and the daily (level 1) carbohydrate consumption classification. Therefore, non-parametric MLCA was employed 1) to identify latent classes of observation days (level 1) based on the subjects' responses to the 4-day food and drink diary and 2) to form distinct latent classes of individuals (level 2) based on the distribution of day-level carbohydrate eating temporal latent classes within individuals.\vspace{-0.3cm}

@@

The number of classes in level 1 was determined by 1) the evaluation of model fit indices, including the Bayesian information criterion (BIC) and entropy, which is a statistic that summarizes latent class probabilities where values near 1 indicate better latent class separation; 2) the Lo-Mendell-Rubin Likelihood Ratio Test (LMR-LRT) \parencite{lo2001testing, nylund2007deciding} which compares $q$ vs $q-1$ classes models, where $q$ is the number of latent classes; most importantly, 3) pattern interpretability. In the steps of performing multilevel LCA, where LMR-LRT is not available, the same rules of model fit indices and pattern interpretability were used to determine the optimal combination of latent classes in observation day-level and individual-level. MLCA models were fitted in Mplus 7.4 \parencite{muthen2005mplus}, the Mplus syntax and outputs are shown in \textbf{Appendix \ref{AppendixB}}. \vspace{-0.3cm}

@@

Day-level latent classes identified by the first step of MLCA were tabulated according to the day of the week and also whether the diary was recorded during weekends or not. A contingency table giving the frequency of responses across the 7-time slots of the survey days was produced. Descriptive statistics for the dietary day-level recordings according to the latent class memberships were presented. Pearson $\chi^2$ test was used to assess evidence fora difference in the distribution of categorical variables. One-way Analysis of Variance (ANOVA) was used to compare the means across the multiple groups for continuous variables.

@@

Associations between individual-level carbohydrate eating classes and hypertension (yes/no), body mass index (BMI, kg/m\textsuperscript{2}), and waist circumference (WC, cm) were explored in men and women separately. Point estimates of weighted means and proportions and 95\%CI of the characteristics were determined by applying either nurse visiting weights (for outcomes of hypertension, BMI, and WC) or blood sample weights (for diagnosis of DM) accordingly. Similarly, \textit{F} tests (for continuous variables) and adjusted Pearson $\chi^2$ tests (for categorical variables) were used to determine sex-specific differences by hypertension status and BMI categories.

@@

\begin{enumerate}
	\item The crude association between the carbohydrate eating groups and the outcomes was first examined. 
	\item Potential confounders of the association between carbohydrate eating groups (exposure) and the outcomes were selected depending on the descriptive statistical analyses conducted above, i.e. those associated with both the exposure and the outcome and also not on the causal pathway were selected as potential confounders. Covariates that are strongly related with the outcomes but not associate with carbohydrate eating groups may reduce the standard errors and improve the precisions, so they were also considered in the linear regression models. 
	\item Confounding or interaction effects from each of the potential factors were checked one by one. Interaction effects were tested using the adjusted Wald test for whether the regression coefficients of the interaction terms are simultaneously equal to zero.  
	\item A preliminary model that includes all of the variables suggested to be confounders in the previous step was established. 
	\item The remaining variables were added to the preliminary model one by one to see if any of them may be a confounder in the presence of the other covariates. 
	\item For logistic regression models (hypertension) under the survey data, goodness-of-fit was assessed using \textit{F}-adjusted mean residual test \parencite{archer2001goodness,archer2007goodness} which is implemented in Stata using command \code{svylogitgof} \parencite{archer2006goodness}. Other diagnostics for regular logistic regression models, such as estimating the pseudo-R\textsuperscript{2},  AIC or BIC, checking the standardised Pearson residuals, or covariate pattern residuals are currently not available for weighted survey data. 
	\item For linear regression models (WC, BMI), the assumption of independent observations is violated because of the clustered structure of the data. Therefore, inference from such models is based on design rather than the model.  
	General checking such as QQ plots of the residuals (normality), plotting the residuals against fitted values (constant variance) are unneeded (also not available). Outliers, leverage, and Cook's distance cannot be checked either. However, participants with extreme weightings (if any) were checked by removing them and refitting the models as a sensitivity analysis.
	\item Since under complex survey designed data, the sampling-weighted least squares are not maximum likelihood, it would not be possible to compare models using likelihood ratio test. Instead, adjusted Wald tests with $p < 0.05$ were used as criteria for variable inclusion in the final model. Another Stata command \code{linktest} was used to decide whether quadratic and cubic terms of continuous variables were necessary for improving the fitting of the model \parencite{pregibon1980goodness}. 
\end{enumerate}
